Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-09-24 paper/paper.tex
directives ofr result section
refactor introduction

% Section outline:
% - How to accurately model timbre space? Useful for studying auditory perception, browsing databases, etc.
% - Studies either consider stimuli that are ``too simple'' (e.g., tones, instruments with fixed playing techniques) or ``too complex'' (e.g., speech, environmental sounds).
% - Computational models do not capture enough information (MFCCs and related representations) or are hard to compute or analyze (STRFs, deep networks).
% - Present a study of timbral similarity perception on diverse set of instrumental sounds with extended playing techniques (PTs). Provides more control in modeling different aspects of timbre.
% - In addition, propose a computational model for timbral similarity: scattering (separable or joint) plus a linear layer. Fixed representation to extract time-frequency structure with adaptivity in linear component.

One of the most important properties for identifying the source of a sound is its \emph{timbre}.
While it is typically defined as the quality of a sound that is independent of its pitch, duration, or loudness, this definition fails to provide an explicit characterization of the attribute.
Due to its central role in audio perception, timbre has been studied from several angles, including perceptual and psychophysical studies.
Since timbre extraction forms the basis of many audio-based information retrieval tasks, such as classification and determining acoustical similarity, the subject has also been studied from a machine listening perspective.
\ja{Need references for the above statements.}

A natural candidate for characterizing timbre is the spectral content of a signal.
This information, however, is not rich enough to fully characterize the timbre of a sound, as evidenced by the recent move from spectral descriptors such as mel-spectrograms and mel-frequency cepstral coefficients (MFCCs) and descriptors of spectral energy distributions.
Indeed, the success of deep neural networks (DNNs) applied to time--frequency representations (such as spectrograms or scalograms) in machine listening have demonstrated the necessity for capturing both spectral and temporal content.
In neurophysiological, the auditory perception models proposed by S. Shamma and collaborators, known as \emph{spectro--temporal receptive fields} (STRFs) all characterize temporal structure in addition to purely spectral information.
It seems, therefore, that timbre is inherently spectro--temporal.
\ja{Again, refs.}

While the above models offer some insight into the structure of timbre, their complexity makes interpretation and analysis difficult.
The expressive power of DNNs are due to tuning a large number of parameters, so it is often hard to relate a specific component to a property of the resulting perception model.
STRFs, on the other hand, have a simpler structure, but are ultimately constructed to model electrophysiological signal in the auditory cortex.
As a result, they consist several adjustments that make their computation costly and their mathematical analysis difficult.
\ja{Refs.}

Another difficulty in studying the perception of timbre is the choice of stimuli.
Previous work on timbre characterization of STRFs \cite{patil2012music} has focused on instrumental sounds of isolated notes played with standard playing techniques.
This inherently restricts the range of available timbre studied, limiting the conclusions of the work.
On the other end of the spectrum are computational models for information retrieval tasks.
Since these are primarily concerned with performance on real-world sounds, the possible range of timbre is very broad.
The analysis of the resulting model is therefore all the more difficult, since we have little control over the complexity of the stimuli.
\ja{Refs.}

Our contribution is twofold.
First, we propose a simple model for timbre perception involving a fixed time--frequency descriptor, the scattering, followed by a learned linear layer optimized for the timbral space under consideration.
Second, we validate this model on a set of instrumental sounds with extended playing techniques (PTs), allowing for a richer range of timbre compared to ordinary instrumental sounds while retaining the structure of the space necessary for interpretability.

As part of validating the model, we have collected timbral similarity judgments in a crowdsourced listening experiment.
These are used as a ``ground truth'' dataset, which allow us to evaluate the model's ability to characterize the perceived similarity of the stimuli.
Analysis of these experimental results also confirms the necessity of including a range of PTs, as these are shown to influence similarity judgments to the same extent as the identity of the instrument itself.

We conclude with a discussion of potential applications to browsing interfaces for musical databases, where composers can easily find sounds on a query-by-example basis.
The adaptivity of the proposed model has the additional advantag of allowing it to be adjusted to a particular listener's preferences by retraining the linear layer.

@@

>>>>

Music information retrieval operates at either of two levels: symbolic and auditory.
By resting on a notation system, the symbolic level facilitates the comparison of musical notes in terms of quantitative attributes, such as duration, pitch, or intensity at the source.
Timbre, in contrast, is a qualitative attribute of music, thus irreducible to a one-dimensional axis.
As a result, symbolic representations describe timbre indirectly, either via visuotactile metaphors (\eg{} dry, rough, bright, round, and so forth) or by specifying an instrumental playing technique (\eg{} bowed or plucked).

Despite their widespread use, linguistic references to musical timbre suffer from limited adequacy.
On the one hand, adjectives such as \emph{bright} or \emph{rough} are prone to misunderstanding, as they do not prescribe any musical gesture that is capable of achieving them.
On the other hand, the sole mention of playing technique does not determine any intended effect; and therefore, fails to generalize to the entire instrumentarium.
For instance, although the term \emph{breathy} alludes to a playing technique that is peculiar to wind instruments, a cellist may accomplish a seemingly breathy timbre by bowing near the fingerboard, i.e., \emph{sul tasto} in Italian.

The prospect of retrieving timbre perception across playing techniques necessarily exceeds the symbolic domain.
Instead, it involves a cognitive process which arises from the subjective experience of listening to music.
From the standpoint of audio signal processing, the task of simulating this cognitive process amounts to the design of a multidimensional feature space wherein some distance function evaluates arbitrary pairs of stimuli.
Rather than to merely discriminate instruments as mutually exclusive categories, this distance function must reflect a consensus of quantitative judgments of acoustic dissimilarity, all other parameters --- \ie{} duration, pitch, and intensity --- being equal.

Behind the overarching challenge of coming up with a robust predictive challenge for human behavior, the main practical application of timbre similarity retrieval lies in the emerging topic of computer-assisted instrumentation.
Following the aesthetic tradition of spectralism in contemporary music creation, the composer queries the machine with an arbitrary audio signal.
The outcome of this is an instrumental audio sample, alongside textual attributes of playing technique.
In this context, the computer serves as a bridge from the auditory level to the symbolic level, allowing to approximate a potentially infinite realm of timbral sensations by an orchestra of priorly specified extent and technical range.
To maximize the perceptual fidelity of such an approximation, a query-by-example search engine in the auditory domain should summarize the vastest possible vocabulary of timbral templates.

In this paper, we propose a machine listening model which computes the dissimilarity in timbre between any two audio samples.
The main originality of our model is that it encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
A second originality of our model is that we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.

%% TODO: list Sections

%% TODO: mention companion website and open-source software

@@

post processing

std consisten across features
mfcc cut
consistent with [previous papers] each item provides incremental gain of perforamnces

table 3 valeurs erreur

ablation study

tfscat+lmnn
tfscat
scat

9 pas de weighting averaging not end to end

separate pas trop de perte : consensus bon

perf du best tool chain on complete dataset

randomize

perf on 50/50

lda

-lmnn

tf -> scat

tfscat -> mfcc + etendus
